---
title: "Reproducible Research: Peer Assessment 1"
author: "James Kowalik"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
I will load the data by first downloading the zip file from the URL, then unzipping, and then storing in the object `data`. 
```{r}
URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(URL, "./repdata%2Fdata%2Factivity.zip")
unzip("./repdata%2Fdata%2Factivity.zip", exdir = ".")
data <- read.csv("./activity.csv")
```

## What is mean total number of steps taken per day?
I will calculate the total number of steps taken per day, first loading `tidyverse` so that `ggplot` can be used and also removing na values.
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```
```{r}
total_steps <- data %>% 
  group_by(date) %>% 
  summarise(total_steps_per_day = sum(steps)) %>%
  .[!is.na(.$total_steps_per_day),] %>%
  print(n = Inf)
```

I will print a histogram of the total steps per day.
```{r}
total_steps %>%
  ggplot(aes(total_steps_per_day)) +
  geom_histogram(binwidth = 2000, col = "black", fill = "blue") + 
  ggtitle("Histogram of Total Steps Per Day") +
  xlab("Total Steps Per Day")
```

I will print the mean and median, converting to a data frame in the final step to show the true mean without rounding.
```{r}
total_steps %>% summarise(mean = mean(total_steps_per_day), median = median(total_steps_per_day)) %>%
  as.data.frame()
```

## What is the average daily activity pattern?
I will create a time series plot of the average daily steps taken per 5 minute time interval, first creating the object `average_steps` to be used both now and later, then removing na values in calculating the mean, and finally plotting with an adjustment to the title font size.
```{r}
average_steps <- data %>% 
  group_by(interval) %>%
  summarise(mean = mean(steps, na.rm = TRUE))
average_steps %>% ggplot(aes(interval, mean)) +
  geom_line(col = "blue") +
  ggtitle("Time Series Plot of the Average Number of Steps for Each 5-minute Interval") +
  theme(plot.title = element_text(size=10.5))
```

Next, I will use the `average_steps` object to print the interval with the maximum average. 
```{r}
average_steps %>% .[which.max(.$mean),] %>%
  .$interval
```

## Imputing missing values
I will print the total number of missing values.
```{r}
sum(is.na(data))
```

I will replace all missing values with the average value (across all dates) for that interval. I have already stored these average values in the object `average_steps`. I will then create a new data set, stored as `data2`, where I merge with the averages, replace the missing values and reorder the columns so as to obtain a dataset in the same form as the original dataset, `data`.
```{r}
merged_data <- merge(data, average_steps)
merged_data$steps[is.na(merged_data$steps)] <- merged_data$mean[is.na(merged_data$steps)]
data2 <- merged_data[order(merged_data$date),]
data2 <- data2[,c(2,3,1)]
```

I will repeat the process of my answer to the first question for `data2`. First, the table.
```{r}
total_steps2 <- data2 %>% 
  group_by(date) %>% 
  summarise(total_steps_per_day = sum(steps)) %>%
  print(n = Inf)
```

Second, the histogram.
```{r}
total_steps2 %>%
  ggplot(aes(total_steps_per_day)) +
  geom_histogram(binwidth = 2000, col = "black", fill = "blue") + 
  ggtitle("Histogram of Total Steps Per Day, NA values Amended") +
  xlab("Total Steps Per Day")
```

Finally, the mean and median.
```{r}
total_steps2 %>% summarise(mean = mean(total_steps_per_day), median = median(total_steps_per_day)) %>%
  as.data.frame()
```
Imputing the missing values has caused no change in the mean but a slight increase in the median- it is in fact now the same as the mean.

## Are there differences in activity patterns between weekdays and weekends?
I will add a column to `data2` indicating whether the day was a weekday or a weekend day by first converting the date variable to date format and then adding it to create a new object `data3`.
```{r}
data2$date <- as.Date(data2$date, format = "%Y-%m-%d")
data3 <- data2 %>% mutate(Type_of_Day = ifelse(weekdays(date) == "Saturday" | weekdays(date) == "Sunday", "Weekend", "Weekday"))
```

I will finally make a panel plot by amending the code written for the second question.
```{r, message = FALSE}
average_steps3 <- data3 %>% 
  group_by(interval, Type_of_Day) %>%
  summarise(mean = mean(steps, na.rm = TRUE))
average_steps3 %>% ggplot(aes(interval, mean)) +
  geom_line(col = "blue") +
  ggtitle("Time Series Plot of the Average Number of Steps for Each 5-minute Interval for Weekdays and Weekends") +
  theme(plot.title = element_text(size=9)) +
  facet_grid(Type_of_Day ~ .)
```
